---
title: "Transition Matrices"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{blkarray}
---


Transition matrix from one location for early summer subseason fire risk: 



\begin{align}
P = 
\begin{blockarray}{cccccc}
& \mbox{Nil} &  \mbox{Low} &\mbox{Moderate}  & \mbox{High} & \mbox{Extreme} \\ 
\begin{block}{c(ccccc)}
\mbox{Nil} & 0.575 & 0.118 & 0.172 & 0.109 & 0.026 \\
\mbox{Low} & 0.453 & 0.243 & 0.148 & 0.123 & 0.033 \\
\mbox{Moderate} & 0.104 & 0.343 & 0.367 & 0.167 & 0.019 \\
\mbox{High} & 0.015 & 0.066 & 0.318 & 0.505 & 0.096 \\
\mbox{Extreme} & 0.000 & 0.060 & 0.149 & 0.567 & 0.224 \\
\end{block}
\end{blockarray}
\end{align}


```{r}
weather_P = matrix(c(0.575,0.118 , 0.172 , 0.109 , 0.026, 
0.453 , 0.243 , 0.148 , 0.123 , 0.033, 
0.104 , 0.343 , 0.367 , 0.167 , 0.019 ,
0.015 , 0.066 , 0.318 , 0.505 , 0.096 , 
0.000 , 0.060 , 0.149 , 0.567 , 0.224),ncol=5,nrow=5,byrow=T)
```

### Code for computing $P^n$ from textbook

```{r}
##### Matrix powers ###############################
# matrixpower(mat,k) mat^k
#
matrixpower = function(X, n){
  
  if(dim(X)[1] != dim(X)[2]){
     throw("Dimensions of the matrix do not match: ", dim(X))
  }
  if (n==0){return(diag(dim(X)[1]))}
  else if(n >0) {return(X %*% matrixpower(X, (n-1)))}
  else {return(matrixpower(solve(X), -n))}
}
```

### For n = 3
```{r}
matrixpower(weather_P,3)
```


### For n = 10
```{r}
matrixpower(weather_P,10)
```

### Simulating from the chain (Code from textbook)

```{r}
###### Simulate discrete-time Markov chain ########################
# Simulates n steps of a Markov chain 
# markov(init,mat,n,states)
# Generates X0, ..., Xn for a Markov chain with initiial
#  distribution init and transition matrix mat
# Labels can be a character vector of states; default is 1, .... k

markov <- function(init,mat,n,labels) { 
	if (missing(labels)) labels <- 1:length(init)
simlist <- numeric(n+1)
states <- 1:length(init)
simlist[1] <- sample(states,1,prob=init)
for (i in 2:(n+1)) 
	{ simlist[i] <- sample(states,1,prob=mat[simlist[i-1],]) }
labels[simlist]
}
####################################################
```

```{r}
library(tidyverse)
###@ Define initial distribution 

init_prob = c(0.4,0.25,0.2,0.1) ## First four states
init_prob = c(init_prob, 1 - sum(init_prob)) ## Forces sum to one

iter=100
simulated_states = markov(init_prob,weather_P,iter,
                          labels=c("Nil","Low","Moderate","High","Extreme"))

simulated_states  = ordered(simulated_states, 
                            c("Nil","Low","Moderate","High","Extreme"))

sim_df = data_frame(Iter=0:iter, States = simulated_states)
ggplot(sim_df,aes(x=Iter,y=States,group=1)) + 
  geom_point(aes(colour=States)) +geom_line(aes(colour=States))

sim_df %>% group_by(States) %>% summarise(Counts=n())
```

### Larger number of iterations 

```{r}
init_prob = c(0.4,0.25,0.2,0.1) ## First four states
init_prob = c(init_prob, 1 - sum(init_prob)) ## Forces sum to one

iter=10000
simulated_states = markov(init_prob,weather_P,iter,
                          labels=c("Nil","Low","Moderate","High","Extreme"))

sim_df = data_frame(Iter=0:iter, States = 
                      ordered(simulated_states,
                              c("Nil","Low","Moderate","High","Extreme")))

ggplot(sim_df,aes(x=Iter,y=States,group=1)) + 
  geom_point(aes(colour=States)) +geom_line(aes(colour=States))

sim_df %>% group_by(States) %>% summarise(Counts=n()) %>% 
    mutate(Proportion=Counts/sum(Counts))

```


